Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-06-24 kappa.tex
Add surface mass balance methods.

@@

\begin{equation}
  \phi(x,y) =
  \begin{cases}
    \phi_0, & \mathrm{for}\ b \le b_0, \\
    \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0},
        & \mathrm{for}\ b_0 < b < b_1, \\
    \phi_1, & \mathrm{for}\ b_1 \le b.
  \end{cases}
  \label{eq:phipiecewise}
\end{equation}

@@

where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.
%$\nu=1\times10^{21}$\,Pa\,s
%$D=5\times10^{24}$\,N\,m
%$\rho_r = 3300$\,kg\,m$^{-3}$

@@

>>>>

% Common units
\newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
\newcommand{\degC}[0]{\unit{{\degree}C}}

@@

\begin{equation}
    \phi(x,y) =
    \begin{cases}
        \phi_0, & \mathrm{for}\ b \le b_0, \\
        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0},
            & \mathrm{for}\ b_0 < b < b_1, \\
        \phi_1, & \mathrm{for}\ b_1 \le b.
    \end{cases}
\end{equation}

@@

where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.
%$\nu=1\times10^{21}$\,Pa\,s
%$D=5\times10^{24}$\,N\,m
%$\rho_r = 3300$\,kg\,m$^{-3}$

@@

\newcommand{\PDD}[0]{\mathrm{PDD}}
\newcommand{\sPDD}[0]{\sigma_{\mathrm{PDD}}}

At the interface between the ice sheet and the atmosphere (Fig.~??),
surfice mass fluxes are computed from monthly mean surface air temperature,
$T_m$, and monthly precipitation, $P_m$, by a temperature-index model
\citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
with temperature between $T_0$ and ${T_1=2\,\degC}$,

\begin{equation}
    a_s =
    \begin{cases}
        0,      & \mathrm{for}\ T_m \le T_0, \\
        P_m \frac{T_m-T_0}{T_1-T_0},
                & \mathrm{for}\ T_0 < b < T_1, \\
        P_m,    & \mathrm{for}\ T_1 \le T_m,
    \end{cases}
\end{equation}

Surface melt, $m_s$, is assumed proportional to the number of positive degree
days (PDD), defined over an arbitrary time interval, $[t_1, t_2]$, as the
integral of positive Celcius temperature $T-T_0$,

\begin{equation}
    \mathrm{PDD} = \int_{0}^{A}\max(T(t)-T_0,0)\,dt.
\end{equation}

For multimillenial simulations of palaeo-ice sheet evolution such as the ones
presented in this thesis, daily or hourly temperature data is not available,
forcing the PDD computation to rely on an idealised representation of the
annual temperature cycle $T_{ac}$. Sub-annual temperature variability around
the freezing point, however, significantly affects surface melt on a multi-year
scale \citep{Arnold.Mackay.1964}. It is then typically included in the models
under an assumption of normal temperature distribution, using a standard
deviation parameter $\sPDD$ in the PDD computation \citep{Braithwaite.1984},
PDDs can then be computed using a double-integral formulation
\citep{Reeh.1991},

\begin{equation}
    \PDD = \frac{1}{\sPDD\sqrt{2\pi}}
        \int_{t_1}^{t_2} \mathrm{d}t
        \int_{0}^{\infty} \mathrm{d}T \,
        T \exp\left({-\frac{(T-T_{ac})^2}{2\sPDD^2}}\right)
\end{equation}

or more efficiently using an error function formulation
\citep{Calov.Greve.2005},

\begin{equation}
    \label{eqn:calovgreve}
    \PDD = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sPDD}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{ac}^2}{2\sPDD^2}}\right)
              + \frac{T_{ac}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{ac}}{\sqrt{2}\sPDD}\right)\right],
\end{equation}

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, surface melt
is computed from PDDs and the available snow and ice depths using an algorithm
documented in the on-line documentation of PISM \citep{PISM-authors.2014}. This
algorythms employ different melt factors for snow ($F_s$) and for ice ($F_i$),
as derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
Paper~I, the temperature standard deviation $\sPDD$ is a constant model
parameter. In Paper~IV, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in Paper~II, and further improved in Paper~III.
%3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
%4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice
